---
title: "Fuel Efficiency and Engine Size"
author: "Clara Reckhorn"
format: 
  closeread-html:
    theme: cosmo
    cr-style:
      section-background-color: white
      narrative-background-color-sidebar: black
      narrative-text-color-sidebar: white
      narrative-font-size: 1.8em
embed-resources: true
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
```

:::: {.cr-section}

How Engine Size Shapes Fuel Efficiency

Modern cars vary widely in fuel efficiency.
As you scroll, we’ll explore how engine displacement relates to miles per gallon.

Step 1 — The Full Dataset

::: {.cr-trigger data-target="cr-plot1"}

At first glance, fuel efficiency might look random across vehicles.

Each point represents a car model in the mpg dataset.

Scroll to investigate whether a pattern begins to emerge.

:::

Step 2 — A Relationship Appears

::: {.cr-trigger data-target="cr-plot1"}

Now focus on the overall direction of the cloud of points.

Do you notice how fuel efficiency changes as engine size increases?

Larger engines generally consume more fuel.

:::

Step 3 — Interpreting the Trend

::: {.cr-trigger data-target="cr-plot1"}

The downward trend suggests a negative relationship:

Higher displacement → lower MPG

Smaller engines → higher efficiency

This is a classic tradeoff between performance and efficiency.

:::

::::

Sticky Visualization

::: {#cr-plot1}

```{r}

ggplot(mpg,
       aes(x = displ, y = hwy)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Engine Displacement vs Highway MPG",
    x = "Engine Displacement (Liters)",
    y = "Highway Miles per Gallon"
  ) +
  theme_minimal(base_size = 16)
```

:::

