---
title: "APIs II"
subtitle: "Web Technologies"
format: stat133slides-revealjs
revealjs-plugins:
  - drop
chalkboard: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE)
library(tidyverse)
library(jsonlite)
library(httr2)
library(leaflet)
```


##

```{r}
#| eval: false
# packages
library(tidyverse)   # ecosystem of data science packages

library(jsonlite)    # JSON parser and generator

library(httr2)       # perform HTTP requests and process responses

library(leaflet)     # web interactive maps
```


# Example: BART API

##

BART Legacy API

<https://www.bart.gov/schedules/developers/api>

- Like many other APIs, **_BART Legacy API_** requires an [API key]{.hilit}.

- Usually, you would have to register with the API provider to obtain an API key.

- Good news: BART lets you use their _Legacy API key_ without having to register.

- [Legacy API key]{.bold-hilit}: `MW9S-E7SL-26DU-VV8V`

- For those interested in getting their own API key, BART's API also provides this kind of key.


## API's Documentation

- In order to use an API, the first step consists of finding and reading the [API's Documentation]{.bold-hilit}.

- The Documentation defines what information you can ask for and how to format your request.

- Keep in mind that some API's provide minimal or almost no documentation, while others provide extensive---and sometimes not-user-friendly---documentation.

- BART Legacy API provides a fair amount of documentation but it can feel a bit intimidating to a beginner user.


## {background-image="images/bart-api1.png" background-size="contain"}

## {background-image="images/bart-api2.png" background-size="contain"}

## {background-image="images/bart-api3.png" background-size="contain"}

## {background-image="images/bart-api4.png" background-size="contain"}

## {background-image="images/bart-api5.png" background-size="contain"}

## {background-image="images/bart-api-query1.svg" background-size="contain"}

## {background-image="images/bart-api-query2.svg" background-size="contain"}

## {background-image="images/bart-api-query3.svg" background-size="contain"}

## {background-image="images/bart-api6.png" background-size="contain"}

## Command: `stns`

API Documentation: <https://api.bart.gov/docs/stn/stns.aspx>

. . .

\

Parameters: 

- `cmd=stns`

- `key=<key>`

- `json=y`

. . .

```
"https://api.bart.gov/api/stn.aspx?cmd=stns&key=INSERT-KEY&json=y"
```


## Interacting from R {auto-animate=true}

. . .

```{r}
#| eval: false
#| code-line-numbers: "1-2|3|1-4"
# set of all available stations
api_url <- "https://api.bart.gov/api/stn.aspx"
res <- request(api_url)
res
```

. . .

```
<httr2_request>
GET https://api.bart.gov/api/stn.aspx
Body: empty
```

## Interacting from R {auto-animate=true}

```{r}
#| eval: false
#| code-line-numbers: "1-3|4|5|6|7|1-8"
# set of all available stations
api_url <- "https://api.bart.gov/api/stn.aspx"
res <- request(api_url) |>
  req_url_query(cmd  = "stns") |>
  req_url_query(key  = "MW9S-E7SL-26DU-VV8V") |>
  req_url_query(json = "y") |>
  req_perform()
res
```

. . .

```
<httr2_response>
GET https://api.bart.gov/api/stn.aspx?cmd=stns&key=MW9S-E7SL-26DU-VV8V&json=y
Status: 200 OK
Content-Type: application/json
Body: In memory (10289 bytes)
```


## Interacting from R {auto-animate=true}

```{r}
#| echo: false
# set of all available stations
api_url <- "https://api.bart.gov/api/stn.aspx"
res <- request(api_url) |>
  req_url_query(cmd  = "stns") |>
  req_url_query(key  = "MW9S-E7SL-26DU-VV8V") |>
  req_url_query(json = "y") |>
  req_perform()
res
```

```{r}
data <- resp_body_json(res, simplifyVector = FALSE)
names(data)
```

. . .

```{r}
names(data$root)
```

. . .

```{r}
names(data$root$stations)
```

. . .

```{r}
# number of stations
length(data$root$stations$station)
```


##

:::: {.columns}
::: {.column}
```{r}
data$root$stations$station[[1]]
```
:::

::: {.column}
```{r}
data$root$stations$station[[2]]
```
:::

::::


##

```{r}
#| output-location: fragment
# extract station names (into a vector)
stn_names <- sapply(data$root$stations$station, function(x) x$name)
head(stn_names)
```

. . .

```{r}
#| output-location: fragment
# extract station latitude and longitude coords
stn_lat <- sapply(data$root$stations$station, 
                  function(x) as.numeric(x$gtfs_latitude))

stn_lon <- sapply(data$root$stations$station, 
                  function(x) as.numeric(x$gtfs_longitude))
```


## {auto-animate=true}

```{r}
#| eval: false
# station coords
stn_tbl <- tibble(
  name = stn_names, 
  lat = stn_lat,
  lon = stn_lon)
```

## {auto-animate=true}

```{r}
#| output-location: fragment
# station coords
stn_tbl <- tibble(
  name = stn_names, 
  lat = stn_lat,
  lon = stn_lon)

stn_tbl |> 
  leaflet() |> 
  addTiles() |> 
  addCircleMarkers(label = ~name)
```


## {background-image="images/bart-api7.png" background-size="contain"}

## Command: `stninfo`

API Documentation: <https://api.bart.gov/docs/stn/stninfo.aspx>

. . .

\

Parameters: 

- `cmd=stninfo`

- `key=<key>`

- `orig=<station>`

- `json=y`

. . .

```
"https://api.bart.gov/api/stn.aspx?cmd=stninfo&orig=DBRK&key=INSERT-KEY&json=y"
```


## Detailed information about the specified station

```{r}
# detailed information about the specified station
api_url <- "https://api.bart.gov/api/stn.aspx"
res2 <- request(api_url) |>
  req_url_query(cmd  = "stninfo") |>
  req_url_query(orig = "DBRK") |>
  req_url_query(key  = "MW9S-E7SL-26DU-VV8V") |>
  req_url_query(json = "y") |>
  req_perform()
res2
```

. . .

```
<httr2_response>
GET
https://api.bart.gov/api/stn.aspx?cmd=stninfo&orig=DBRK&key=MW9S-E7SL-26DU-VV8V&json=y
Status: 200 OK
Content-Type: application/json
Body: In memory (1810 bytes)
```


## Interacting from R {auto-animate=true}

```{r}
downtown_berk <- resp_body_json(res2, simplifyVector = FALSE)
downtown_berk
```

. . .

```{r}
lat <- downtown_berk$root$stations$station$gtfs_latitude
lat
```

