---
title: "Working with Vectors"
format: stat133slides-revealjs
chalkboard: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE)
```

## Review

Last time ...

## Announcements

-   Working locally

-   Office Hours / Group Tutoring

-   Problem Sets

# Vectors

## Vectors

-   Central data structure in R

-   Ordered set of elements

-   **Atomic**: elements of the same type

## Atomic\*

Special values in an atomic vector.

-   `NA`: Not Available / Missing

-   `NULL`: Does not exist (nothing)

-   `NaN`: Not a Number (e.g. dividing by zero)

-   `Inf`: Infinity

## A comment about comments {auto-animate="true"}

Comment character: special character to skip evaluation of remaining characters on a line. R (and Python) use `#`.

::: fragment
```{r}
#| output-location: fragment
# create vector
c("A", "C", "D") # vector created!
```
:::

::: fragment
vs.

```{r}
#| output-location: fragment
create vector
c("A", "C", "D") vector created!
```
:::

# Behavior of Vectors

## Question

Turn to neighbor and speculate: what will each of the following lines of code return?

```{r}
#| eval: false
c(1, 2, 3) + c(2, 4, 8)

c(1, 2, 3) + c(2, 4)

c(1, "dog", TRUE)
```

```{r}
#| echo: false
countdown::countdown(2.5, bottom = 0)
```

## Vectorization

[Vectorized]{.bold-hilit}: Operations and functions that can run on an entire vector at once.

```{r}
dbl_vec <- c(1, 2.3, 4)

log(dbl_vec) # vectorized
```

\

```{r}
sum(dbl_vec) # vector-as-argument
```

## Vector Recycling

[Vector Recycling]{.bold-hilit}: If a function requires a longer vector than provided, it will fill the vector with the existing elements, in order.

```{r}
c(2, 6, 1)^2
```

## Coercion

[Implicit Coercion]{.bold-hilit}: Automatic change of data types to ensure atomic vectors. Changes follow a hierarchy.

logical \< integer \< double \< character

## Coercion

[Explicit Coercion]{.bold-hilit}: R provides functions to explicitly coerce from one type to another.

-   `as.logical()`

-   `as.integer()`

-   `as.double()`

-   `as.character()`

. . .

Depending on what data-type you want to coerce to, these functions may or may not work.

. . .

```{r}
as.double(c(TRUE, FALSE, FALSE, FALSE))
```

# Vector Subsetting

## Square Bracket Syntax

## Subsetting by Index

*aka* numeric subsetting

```{r}
#| output-location: fragment
char_vec <- c("bear", "owl", "squirrel")
char_vec
```

. . .

```{r}
#| output-location: fragment
char_vec[2]
```

. . .

```{r}
#| output-location: fragment
char_vec[2:3] 
```

## Subsetting by Name

*aka* character subsetting

```{r}
#| output-location: fragment
char_vec2 <- c("oski" = "bear", "fiat" = "owl", "lux" = "squirrel")
char_vec2
```

. . .

```{r}
char_vec2["oski"]
```

. . .

```{r}
char_vec2[c("lux", "fiat")]
```

## Subsetting by Logicals

*aka* logical subsetting

```{r}
#| output-location: fragment
char_vec2[c(TRUE, FALSE, TRUE)]
```

. . .

Note: the logical vector should be of the same length.

## Question

Write down three methods to generate each vector from the vector `z`, one for each subsetting method.

```{r}
z <- c(latte = 2, boba = 4, matcha = 6)
```

. . .

Vector 1:

```{r}
#| echo: false
z[2:1]
```

. . .

Vector 2:

```{r}
#| echo: false
z[c(3, 3)]
```

```{r}
#| echo: false
countdown::countdown(2.5, bottom = 0)
```

## Change in place syntax

You can assign into a subsetted vector to change its contents in place.

```{r}
#| output-location: fragment
z <- c(latte = 2, boba = 4, matcha = 6)
z[c(TRUE, FALSE, TRUE)] <- 12
z
```

. . .

```{r}
#| output-location: fragment
names(z)
```

. . .

```{r}
#| output-location: fragment
names(z) <- c("A", "B", "C")
z
```

## For next time

-   Work on problem sets.

-   Visit office hours and group tutoring.

-   Have a great weekend 😎
