---
title: "Choropleth Maps"
subtitle: "Maps of Raster Data"
format: stat133slides-revealjs
chalkboard: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE)
library(tidyverse)
library(rvest)
library(sf)
library(viridis)
library(usmap)
library(RColorBrewer)
library(ggiraph)
```

##

```{r}
#| eval: false
# packages
library(tidyverse)     # ecosystem of data science packages
library(rvest)         # for web scraping
library(sf)            # "simple features" (classes for maps)
library(usmap)         # map of US states (with Alaska & Hawaii)
library(RColorBrewer)  # Brewer color palettes
library(ggiraph)       # make ggplots interactive
```


# Average Salary by State (2025)

## {background-image="images/average-salary-by-state-in-the-us-2025.png" background-size="contain"}

## Average Salary by State in 2025

Goal: replicate the choropleth map that appears in the following Forbes's article by Jennifer Simonson (Jan 30, 2026):

\

<https://www.forbes.com/advisor/business/average-salary-by-state/>


## Data

👎 Bad news: _Average Salary by State_ data is displayed in the above Forbes' webpage.
Unfortunately, this HTML table is not scrapable. 

\

👍 Good news: similar data can be found in 
the following SoFi webpage:

<https://www.sofi.com/learn/content/average-salary-in-us/>

\

```{r}
# run this command in the console
# (assuming the working directory is the same of the qmd file)
download.file(
  url = "https://www.sofi.com/learn/content/average-salary-in-us/", 
  destfile = "sofi-average-salary-in-us-2025.html")
```


## {background-image="images/sofi-average-salary-by-state-2025.png" background-size="contain"}


##

Let's extract all the HTML tables from the downloaded HTML file

```{r}
tbls = read_html("sofi-average-salary-in-us-2025.html") |>
  html_table()
```

. . .

\

The table of interest is in the first element of the list.

```{r}
#| output-location: fragment
salaries = tbls[[1]]
head(salaries)
```


##

How would you "convert" `Average` and `Median` to double types?

```{r}
head(salaries)
```

```{r}
#| echo: false
countdown::countdown(1, bottom = 0)
```


## A little bit of data cleaning

```{r}
# data with average and median salaries
salaries = salaries |>
  mutate(Average = as.numeric(str_remove_all(Average, "[,$]")),
         Median = as.numeric(str_remove_all(Median, "[,$]")))

head(salaries)
```


# Map of U.S. States from package `"usmap"`

## Map data with all 50 U.S. States

```{r}
# from package "usmap"
us_states = us_map(regions = "states")

class(us_states)
```

. . .

\

```{r}
names(us_states)
```


## {.smaller}

```{r}
head(us_states)
```

. . .

\

### Columns

+ `fips`: FIPS code, or Federal Information Processing Standard code, 
which is a unique numerical identifier assigned to geographic regions, 
primarily in the United States, by the U.S. federal government.
+ `abbr`: 2-letter abbreviation of State
+ `full`: full name of State
+ `geom`: spatial vector geometry (polygons in this case)


##

We can plot it with `ggplot()` and `geom_sf()`

```{r}
ggplot(data = us_states) +
  geom_sf()
```


# Choropleth Map

## So far: we have 2 tables

. . .

:::: {.columns}
::: {.column}
`salaries`: table of salaries by states with `r nrow(salaries)` rows and the 
following `r ncol(salaries)` columns:

::: {.nonincremental}
+ `State`: full name of State
+ `Average`: Average Salary
+ `Median`: Median Salary
:::
:::

::: {.column .fragment}
`us_states`: table of states and their geometries (this is an `"sf"` object)
with `r nrow(us_states)` rows[^1] and the following `r ncol(us_states)` columns:

::: {.nonincremental}
+ `fips`: FIPS code
+ `abbr`: abbreviated name of State
+ `full`: full name of State
+ `geom`: vector geometry
:::

[^1]: It includes DC
:::
::::


## So far: we have 2 tables {.smaller}

:::: {.columns}
::: {.column width="70%"}
```{r}
us_states
```
:::

::: {.column width="30%"}
```{r}
salaries
```
:::
::::


## Joining `us_states` and `salaries`

```{r}
# inner join
joined_inner = inner_join(
    x = us_states, 
    y = salaries, 
    by = c("full" = "State"))
```


## `inner_join()` {.smaller}

```{r}
# inner join
joined_inner = inner_join(us_states, salaries, by = c("full" = "State"))
dim(joined_inner)
```

\

```{r}
head(joined_inner)
```


## Choropleth Map (ver 1)

```{r}
joined_inner |>
  ggplot() +
  geom_sf(aes(fill = Average), color = "white") +
  theme_void()
```


## Purple Color Palette

To have a color scheme similar to the plot in Forbes' article, we can use
`scale_fill_gradient()`

```{r}
#| eval: false
joined_inner |>
  ggplot() +
  geom_sf(aes(fill = Average), color = "white") +
  scale_fill_gradient(low = "#1b4bab", high = "#d7e1f5") +
  theme_void()
```


##

```{r}
#| code-fold: true
joined_inner |>
  ggplot() +
  geom_sf(aes(fill = Average), color = "white") +
  scale_fill_gradient(low = "#1b4bab", high = "#d7e1f5") +
  theme_void()
```


## Another Purple Color Palette

We can also use `scale_fill_gradient()`. In this case we borrow one of the 
color schemes from the package `"RColorBrewer"`: the `"Purples"` scheme

<https://colorbrewer2.org/#type=sequential&scheme=Purples&n=9>

```{r}
#| eval: false
joined_inner |>
  ggplot() +
  geom_sf(aes(fill = Average), color = "white") +
  scale_fill_gradientn(colors = brewer.pal(n = 7, name = "Purples"), 
                       breaks = seq(45000, 80000, by = 5000)) +
  theme_void()
```

\

In contrast to the map in Forbes, we prefer to have dark hues for higher average
salaries, and light hues for lower average salaries.


##

```{r}
#| code-fold: true
joined_inner |>
  ggplot() +
  geom_sf(aes(fill = Average), color = "white") +
  scale_fill_gradientn(colors = brewer.pal(n = 7, name = "Purples"), 
                       breaks = seq(45000, 80000, by = 5000)) +
  theme_void()
```


# Interactive Map

## Interactive map with `"ggiraph"`

We can make it interactive by using `geom_sf_interactive()` and activating
the `tooltip` argument inside the aesthetic mappings.


##

```{r}
#| code-fold: true
# adding interactivity with "ggiraph" functions
gmap = joined_inner |>
  mutate(Label = paste0(full, ": ", Average)) |>
  ggplot(aes(fill = Average,
             tooltip = Label)) +
  geom_sf_interactive(color = "white") +
  scale_fill_gradientn(colors = brewer.pal(n = 7, name = "Purples"), 
                       breaks = seq(45000, 80000, by = 5000)) +
  theme_void()

girafe(ggobj = gmap)
```


## Viridis Palette

```{r}
#| code-fold: true
# adding interactivity with "ggiraph" functions
gmap = joined_inner |>
  mutate(Label = paste0(full, ": ", Average)) |>
  ggplot(aes(fill = Average,
             tooltip = Label)) +
  geom_sf_interactive(color = "white") +
  scale_fill_viridis_c(direction = -1) +
  theme_void()

girafe(ggobj = gmap)
```

