---
title: "Strings I"
subtitle: "Parsing text with regular expressions"
format: stat133slides-revealjs
chalkboard: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE)
library(tidyverse)
library(rvest)
```


## Agenda

- Fundamentals of character strings
- Regular Expressions
  + Literal Characters
  + Metacharacters
    - Wildcard
    - Escape
  + Character sets
  + Character classes
  + Abbreviated classes


# Fundamentals of Strings

## Character Vectors

```{r}
teas <- c("taro", "green", "black", "matcha", "jasmine")
```

. . .

\

```{r}
vehicles <- c("car", "bike", "train", "boat", "airplane")
```

. . .

\

```{r}
fives <- c("5.00", "5100", "5 00", "5-00", "$5.55")
```


## 

A [character vector]{.hilit} is a set of strings, each of which has a number of characters (`nchar()`).

```{r}
#| output-location: fragment
teas
```

. . .

\

```{r}
#| output-location: fragment
length(teas)
```

. . .

\

```{r}
#| output-location: fragment
nchar(teas)
```


## Creating Strings

::: {.nonincremental}
- Wrap the characters either in `"` or `'`.
- Combine strings with `paste()`
:::

```{r}
#| output-location: fragment
paste("pa", "tro", "num")
```


## Creating Strings

::: {.nonincremental}
- Wrap the characters either in `"` or `'`.
- Combine strings with `paste()`
:::

```{r}
#| output-location: fragment
paste("pa", "tro", "num", sep = ",")
```


## Creating Strings

::: {.nonincremental}
- Wrap the characters either in `"` or `'`.
- Combine strings with `paste()`
:::

```{r}
#| output-location: fragment
paste("pa", "tro", "num", sep = "")
```


## Changing Case

- `tolower()` makes all letter characters lowercase
- `toupper()` makes all letter characters uppercase

. . .

```{r}
#| output-location: fragment
TEAS <- toupper(teas)
TEAS
```

. . . 

\

```{r}
#| output-location: fragment
tolower(TEAS)
```


## Quotations

```{r}
#| output-location: fragment
"Harry pulled out his wand. "Expecto Patronum!" he yelled."
```

. . . 

\

Include a quotation in a string by using `'`.

```{r}
#| output-location: fragment
'Harry pulled out his wand. "Expecto Patronum!" he yelled.'
```


## Some Special Characters

- `"` normally starts or ends a string. `\"` is the double-quote character.

- `t` normally is the lowercase Latin t. `\t` is a tab space.

- `n` normally is the lowercase Latin n. `\n` is a new line.

- U0001F600 normal is just that. `\U0001F600` is [😀.]{.fragment}

```{r}
#| echo: false
#| eval: false
str_view(" I \U0002665\U000FE0F you")
```


# Package `"stringr"`

## `"stringr"`

Provides a large set of functions for string manipulation

:::: {.columns}
::: {.column width="50%"}
- `str_length()`: analog of `nchar()`
- `str_c()`: analog of `paste()`
- `str_view()`
- `str_sub()`
- `str_detect()`
:::

::: {.column width="50%"}
- `str_split()`
- `str_match()`
- `str_replace()`
- `str_extract()`
:::
:::


## `str_view()`

Show more readable version of strings. Also see how a pattern matches.

```{r}
#| output-location: fragment
x <- 'Harry pulled out his wand. "Expecto Patronum!" he yelled.'
x
```

. . .

\

```{r}
#| output-location: fragment
str_view(x)
```


## `str_sub()`

Subset strings by position

. . . 

```{r}
#| output-location: fragment
teas
```

. . . 

\

```{r}
#| output-location: fragment
str_sub(teas, start = 1, end = 2)
```

. . . 

\

```{r}
#| output-location: fragment
str_sub(teas, start = 1, end = 3)
```

. . .

\

```{r}
#| output-location: fragment
str_sub(teas, start = -3)
```


## `str_detect()`

Detects the presence or absence of a pattern in a string

```{r}
teas
```

. . . 

```{r}
#| output-location: fragment
str_detect(teas, pattern = "tar")
```

. . . 

\

```{r}
#| output-location: fragment
str_detect(teas, pattern = "at")
```

. . . 

\

```{r}
#| output-location: fragment
str_view(teas, pattern = "at")
```


# Regular Expressions

## Regular Expressions (regex)

A sequence of characters expressing a string or pattern to be searched for within a longer piece of text.

. . .

Regex is not:

- a programming language
- a markup syntax
- a UNIX utility


## Regex

Simply put, regex is 

a text string (i.e. set of characters) that defines

::: {.fragment .highlight-red}
[a certain amount of text]{.highlight-red}
:::

. . .

\

That "certain amount of text" is called a [pattern]{.hilit}.

. . .

\

Regex, at its core, has to do with matching patterns of text.


## 2 main types of characters

\

::: {.fragment}
### 1) Literal Characters

A literal character is a character that matches itself.
:::

\

::: {.fragment}
### 2) Metacharacters

A metacharacter is a character that does NOT match itself.
:::


## Literal Characters

```{r}
vehicles
```

. . .

\

Match the letter "a".

```{r}
#| output-location: fragment
str_view(vehicles, pattern = "a")
```


## Literal Characters

```{r}
vehicles
```

. . .

\

Match the letter "i".

```{r}
#| output-location: fragment
str_view(vehicles, pattern = "i")
```


## Literal Characters

```{r}
vehicles
```

. . .

\

Match the letters "ai".

```{r}
#| output-location: fragment
str_view(vehicles, pattern = "ai")
```


## Literal Characters

```{r}
vehicles
```

. . .

\

Match the letters "air".

```{r}
#| output-location: fragment
str_view(vehicles, pattern = "air")
```


## Metacharacters

Non-letter characters with special meaning.

- `.`
- `?`
- `+`
- `*`
- `[`
- `]`
- etc


## Wildcard `.`

`.` will match ANY single character

```{r}
fives
```

. . .

\

```{r}
#| output-location: fragment
str_view(fives, pattern = ".")
```


## Wildcard `.`

```{r}
fives
```

. . .

\

5 followed by any character: "5."

```{r}
#| output-location: fragment
str_view(fives, pattern = "5.")
```


## Wildcard `.`

```{r}
fives
```

. . .

\

any character followed by zero: ".0"

```{r}
#| output-location: fragment
str_view(fives, pattern = ".0")
```


## Escape metacharacter `\`

But what if the string has an actual period?

```{r}
fives
```

. . .

\

How do we match `"5."` but not `"51"` (or any other character)?

. . .

```{r}
#| output-location: fragment
str_view(fives, pattern = "5\\.")
```

. . .

The first `\` escapes it for the string, the second `\` escapes it for the regex.


# Character Sets

## Character Set

A character set is a set of characters defined (grouped) within brackets `[  ]`


## Character Set

```{r}
vehicles
```

. . .

\

Match single occurrence of either "a" or "o".

```{r}
#| output-location: fragment
str_view(vehicles, pattern = "[ao]")
```

. . .

[A character set matches any single character in the set]{.mdlit}


## Character Set

```{r}
vehicles
```

. . .

\

Match any vowel.

```{r}
#| output-location: fragment
str_view(vehicles, pattern = "[aeiou]")
```


## Character Set

```{r}
vehicles
```

. . .

\

Match single occurrence of either "k" or "t".

```{r}
#| output-location: fragment
str_view(vehicles, pattern = "[kt]")
```


## Character Set

```{r}
vehicles
```

. . .

\

Match single occurrence of "c", "a", or "r".

```{r}
#| output-location: fragment
str_view(vehicles, pattern = "[car]")
```

. . .

\

BTW: Do NOT confuse `"[car]"` with `"car"`


# Character Ranges

## Character Ranges

A character set is a set of characters defined (grouped) within brackets `[  ]`

- `[0123456789]`
- `[abcdefghijklmnopqrstuvwxyz]`
- `[ABCDEFGHIJKLMNOPQRSTUVWXYZ]`


## Character Ranges

A character set is a set of characters defined (grouped) within brackets `[  ]`

::: {.nonincremental}
- `[0123456789]` $\ \rightarrow \ $ `[0-9]`
- `[abcdefghijklmnopqrstuvwxyz]` $\ \rightarrow \ $ `[a-z]`
- `[ABCDEFGHIJKLMNOPQRSTUVWXYZ]` $\ \rightarrow \ $ `[A-Z]`
:::

- `[012...789abc...xyz]` $\ \rightarrow \ $ `[0-9a-z]`
- `[012...789ABC...XYZ]` $\ \rightarrow \ $ `[0-9A-Z]`
- `[abc...xyzABC...XYZ]` $\ \rightarrow \ $ `[a-zA-Z]`
- `[abcdefg34567UVWXYZ]` $\ \rightarrow \ $ `[a-g3-7U-Z]`

##

```{r}
fives
```

. . .

```{r}
#| output-location: fragment
str_view(fives, pattern = "[0-9]")
```
. . .

```{r}
#| output-location: fragment
str_view(fives, pattern = "[0-9][0-9]")
```

. . .

```{r}
#| output-location: fragment
str_view(fives, pattern = "[0-9][0-9][0-9]")
```


# POSIX Classes

## POSIX Classes

A POSIX class is a special set designed as a standardized set to ensure consistency across different regex tools.

\

POSIX classes are used inside a bracket expression `[ ]` and are themselves enclosed in `[:` and `:]` delimiters.

\

For example `[[:digit:]]` which is equivalent to `[0-9]`


## POSIX Classes

- `[[:digit:]]` analog of `[0-9]`
- `[[:alpha:]]` analog of `[A-Za-z]`
- `[[:alnum:]]` analog of `[A-Za-z0-9a]`
- `[[:lower:]]` analog of `[a-z]`
- `[[:upper:]]` analog of `[A-Z]`
- `[[:space:]]` for whitespace characters
- `[[:punct:]]` for punctuation characters
- `[[:xdigit:]]` for hexadecimal digits
- `[[:blank:]]` for space and tab characters


## 

What will be matched by the following patterns?

```{r}
txt <- c("5\t00", "car", "Bike", "TARO 5.00")
```

```{r}
#| echo: false
countdown::countdown(1.5, bottom = 0)
```

\

. . .

```{r}
#| eval: false
str_view(txt, pattern = "[[:digit:]]")
```

\

```{r}
#| eval: false
str_view(txt, pattern = "[[:space:]]")
```

\

```{r}
#| eval: false
str_view(txt, pattern = "[[:upper:]]")
```


## 

What will be matched by the following patterns?

```{r}
txt <- c("5\t00", "car", "Bike", "TARO 5.00")
```

\

```{r}
str_view(txt, pattern = "[[:digit:]]")
```


```{r}
#| output-location: fragment
str_view(txt, pattern = "[[:space:]]")
```


```{r}
#| output-location: fragment
str_view(txt, pattern = "[[:upper:]]")
```


## 

(cont) What will be matched by the following patterns?

```{r}
txt <- c("5\t00", "car", "Bike", "TARO 5.00")
```

```{r}
#| echo: false
countdown::countdown(1.5, bottom = 0)
```

\

. . .

```{r}
#| eval: false
str_view(txt, pattern = "[[:alnum:]]")
```

\

```{r}
#| eval: false
str_view(txt, pattern = "[[:punct:]]")
```

\

```{r}
#| eval: false
str_view(txt, pattern = "[[:alpha:]]")
```


## 

(cont) What will be matched by the following patterns?

```{r}
txt <- c("5\t00", "car", "Bike", "TARO 5.00")
```

\

```{r}
str_view(txt, pattern = "[[:alnum:]]")
```

```{r}
#| output-location: fragment
str_view(txt, pattern = "[[:punct:]]")
```

```{r}
#| output-location: fragment
str_view(txt, pattern = "[[:alpha:]]")
```


# Abbreviated Classes

## Abbreviated Classes

- `\s` (in R: `\\s`) : any spaces
    + equivalent to `[[:space:]]`

- `\d` (in R: `\\d`) : digits
    + equivalent to `[[:digit]]`
    
- `\w` (in R: `\\w`) : "word" character: lower & upper case letters, digits, \_
    + equivalent to `[A-Za-z0-9_]`

- `\b` (in R: `\\b`) : word boundaries

. . .

There are also negation classes: `\S`, `\D`, `\W`, and `\B` (in R: `\\S`, `\\D`, `\\W`, `\\B`)


##

```{r}
fives
```

. . .

```{r}
#| output-location: fragment
str_view(fives, pattern = "\\d")
```

. . .

```{r}
#| output-location: fragment
str_view(fives, pattern = "\\d\\d")
```


##

```{r}
fives
```

. . .

```{r}
#| output-location: fragment
str_view(fives, pattern = "\\s")
```

. . .

```{r}
#| output-location: fragment
str_view(fives, pattern = "\\S")
```

##

```{r}
fives
```

. . .

```{r}
#| output-location: fragment
str_view(fives, pattern = "\\w")
```

. . .

```{r}
#| output-location: fragment
str_view(fives, pattern = "\\W")
```
