---
title: "Dates and Times II"
subtitle: "Time spans and time zones."
format: stat133slides-revealjs
revealjs-plugins:
  - drop
chalkboard: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE)
library(tidyverse)
library(rvest)
```


## Agenda

Last time: 

1) How does a computer __Store__ dates and times?

2) How do we __Create__ dates and times?

3) How do we __Get and Set__ them?

. . .

Today:

4) How do you specify __Time Spans__?

5) How do you account for __Time Zones__?

. . .

\

```{r}
#| eval: false
library(tidyverse)  # includes "lubridate"
```


# Time Spans

## How old is Oski?

Oski, the official mascot for UC Berkeley, was debuted on September 27, 1941, during a
freshman rally at the Greek Theatre.

. . .

```{r}
#| output-location: fragment
a <- today()
b <- ymd("1941-09-27")
class(a)
```

. . .

```{r}
#| output-location: fragment
class(b)
```

. . .

```{r}
#| output-location: fragment
a - b
```


## Durations

Subtraction of two base R `Date` vectors -> `difftime` vector.

. . .

```{r}
#| output-location: fragment
class(a - b)
```

. . .

A `difftime` can be recorded in many different units, so better to use `lubridate::as.duration()` to always record it in [seconds]{.bold-hilit}.

. . .

```{r}
#| output-location: fragment
oski_age_sec <- as.duration(a - b)
class(oski_age_sec)
```

. . .

```{r}
#| output-location: fragment
oski_age_sec
```


## Constructing a duration

You can make a `duration` by coercing a `Date` object or using a constructor function.

. . .

:::: {.columns}
::: {.column .nonincremental width="50%"}
```{r}
dseconds(15)
```

```{r}
dminutes(10)
```

```{r}
dhours(c(12, 24))
```
:::

::: {.column .fragment .nonincremental width="50%"}
```{r}
ddays(0:3)
```

```{r}
dweeks(3)
```

```{r}
dyears(1)
```
:::
::::


## Some Wrinkles

. . .

```{r}
#| output-location: fragment
one_am <- ymd_hms("2026-03-08 01:00:00", tz = "America/Los_Angeles")
one_am
```

. . .

\

What should the following yield?

```{r}
#| output-location: fragment
one_am + ddays(1)
```

. . .

> Daylight savings time!


## Question

What is the answer to the following questions:

1) How long is a __day__?

2) How long is a __year__?


## Periods

In `"lubridate"`, a `period` is like a `duration` except it doesn’t have a fixed length in seconds. It tracks human units (day, week) that are responsive to irregularities.

. . .

:::: {.columns}
::: {.column .nonincremental width="50%"}
Example period constructors:

```{r}
seconds(15)
```

```{r}
minutes(10)
```

```{r}
hours(c(12, 24))
```
:::

::: {.column .fragment .nonincremental width="50%"}
\
```{r}
days(0:3)
```

```{r}
weeks(3)
```

```{r}
years(1)
```
:::
::::


## Working with periods

. . .

```{r}
#| output-location: fragment
one_am
```

. . .

```{r}
#| output-location: fragment
one_am + days(1)
```

. . .

\

What will the following yield?

```{r}
#| output-location: fragment
ymd("2024-01-01") + dyears(1)
```

. . .

```{r}
#| output-location: fragment
ymd("2024-01-01") + years(1) # accounts for leap year
```


## How old will Oski be ...

. . .

In a year?

```{r}
#| output-location: fragment
oski_age_sec
```

. . .

```{r}
#| output-location: fragment
oski_age_sec + dyears(1) # adding average seconds in a year
```

. . .

```{r}
#| output-location: fragment
as.period(oski_age_sec)
```

. . .

```{r}
#| output-location: fragment
as.period(oski_age_sec) + years(1)
```


## Time Spans recap

### Base R

- `difftime`: inconsistent units

\

### lubridate

- `duration`: fixed span in seconds

- `period`: variable span in human units

- `interval`: duration plus a starting point


# Time Zones

## {background-image="images/World_Time_Zones_Map.png" background-size="contain"}

## {background-image="images/alcan.png" background-size="contain"}

## {background-image="images/india.png" background-size="contain"}

## {background-image="images/china.png" background-size="contain"}


## UTC

Coordinated Universal Time (UTC) is primary global time standard. Successor to GMT.


### Offsets

Each time zone has a particular offset from UTC.

- London: +0
- New York: -5
- Kyoto: +9

::: {.fragment}

```{r}
#| output-location: fragment
Sys.timezone() # What R says my timezone is
```
:::


## Olson Name

```{r}
# names of Time Zones
OlsonNames()
```


## Changing Time

Two flavors

Change the instant in time.


```{r}
#| output-location: fragment
x1 <- ymd_hms("2024-06-01 12:00:00", tz = "America/New_York")
x1
```

. . .

```{r}
#| output-location: fragment
force_tz(x1, tzone = "Asia/Tokyo")
```

. . .

Keep the same instant but change the way it's printed:

```{r}
#| output-location: fragment
with_tz(x1, tzone = "Asia/Tokyo")
```


# Some Examples

## Men's Long Jump World Record Progression {.smaller}

```{r}
#| echo: false
doc <- read_html("mens-long-jump.html")
dat <- html_table(doc)[[2]]
```

```{r}
#| eval: false
library(rvest)
url <- "https://en.wikipedia.org/wiki/Men%27s_long_jump_world_record_progression"

# import HTML file
doc <- read_html(url)

# extract 2nd HTML table (as data.frame/tibble)
dat <- html_table(doc)[[2]]
dat
```

. . .

```{r}
#| echo: false
dat
```


##


```{r}
#| output-location: fragment
#| code-line-numbers: "1-2|1-4|1-6|1-8|1-10"
dat |> 
  mutate(
    # distance
    Mark = as.numeric(str_sub(Mark, start = 1, end = 4)),
    # extract string Date
    Date = str_extract(Date, "\\d+ \\w+ \\d{4}"),
    # convert column  Date as "Date" type
    Date = dmy(Date)) |> 
    select(Mark, Date) |> 
    slice_head(n = 5)
```


##


```{r}
#| code-fold: true
dat |> 
  mutate(
    Mark = as.numeric(str_sub(Mark, start = 1, end = 4)),
    Date = str_extract(Date, "\\d+ \\w+ \\d{4}"),
    Date = dmy(Date)) |> 
  ggplot(aes(x = Date, y = Mark)) +
  geom_point() +
  geom_step() +
  labs(title = "Men's Long Jump World Record Progression") +
  theme_minimal()
```


## 10 Matches of Serena Williams

Data file `serena_williams_matches.csv`

\

```
player1,player1_dob,player2,player2_dob,date_match
Serena Williams,1981-09-26,Iva Majoli,1977-08-12,2001-08-13
Serena Williams,1981-09-26,Maria Kirilenko,1987-01-25,2004-05-24
Serena Williams,1981-09-26,Sania Mirza,1986-11-15,2005-01-17
Serena Williams,1981-09-26,Tatiana Perebiynis,1982-12-15,2007-10-08
Serena Williams,1981-09-26,Maria Jose Martinez,1982-08-12,2009-08-31
Serena Williams,1981-09-26,Maria Jose Martinez,1982-08-12,2010-01-11
Serena Williams,1981-09-26,Caroline Wozniacki,1990-07-11,2014-08-25
Serena Williams,1981-09-26,Venus Williams,1980-06-17,2015-06-29
Serena Williams,1981-09-26,Yaroslava Shvedova,1987-09-12,2016-08-29
Serena Williams,1981-09-26,Maria Sakkari,1995-07-25,2020-08-24
```


## 10 Matches of Serena Williams

Data file `serena_williams_matches.csv`

::: {.nonincremental}
- `player1`: Serena Williams
- `player1_dob`: Serena Williams' date of birth
- `player2`: Opponent's name
- `player2_dob`: Opponent's date of birth
- `date_match`: date of match
:::


## Serena Williams matches {.smaller}

```{r}
matches <- read_csv("serena_williams_matches.csv")
matches
```

. . .

\

1) Compute Serena Williams' age at date of match

2) Compute opponent's age at date of match

3) Compute age difference between Serena and opponent


## Serena's and opponent's age

```{r}
#| output-location: fragment
# Serena's age (as difftime)
matches$date_match - matches$player1_dob
```

. . .

\

```{r}
#| output-location: fragment
# Opponent's age (as difftime)
matches$date_match - matches$player2_dob
```


## Serena's and opponent's age

```{r}
#| output-location: fragment
# Serena's age (as duration)
as.duration(matches$date_match - matches$player1_dob)
```

\

```{r}
#| output-location: fragment
# Opponent's age (as duration)
as.duration(matches$date_match - matches$player2_dob)
```


## Serena's and opponent's age

```{r}
#| output-location: fragment
# Serena's age (as period in "years")
years(as.duration(matches$date_match - matches$player1_dob))
```

\

```{r}
#| output-location: fragment
# Opponent's age (as period in "years")
years(as.duration(matches$date_match - matches$player2_dob))
```


## Serena's and opponent's age

```{r}
#| output-location: fragment
# Serena's age (as interval)
interval(matches$player1_dob, matches$date_match)
```

\

```{r}
#| output-location: fragment
# Opponent's age (as interval)
interval(matches$player2_dob, matches$date_match)
```


## Serena's and opponent's age

```{r}
#| output-location: fragment
# Serena's age (in years)
interval(matches$player1_dob, matches$date_match) / years(1)
```

\

```{r}
#| output-location: fragment
# Opponent's age (in years)
interval(matches$player2_dob, matches$date_match) / years(1)
```


## {.smaller}

```{r}
#| output-location: fragment
matches |> 
  mutate(
    player1_age = interval(player1_dob, date_match) / years(1),
    player2_age = interval(player2_dob, date_match) / years(1),
    age_difference = player1_age - player2_age) |> 
  select(date_match, player1, player1_age, player2, player2_age, age_difference)
```
