---
title: "Dates and Times II"
subtitle: "Time spans and time zones."
format: stat133slides-revealjs
revealjs-plugins:
  - drop
chalkboard: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE)
library(tidyverse)
```


## Agenda

Last time: 

1) How does a computer __Store__ dates and times?

2) How do we __Create__ dates and times?

3) How do we __Get and Set__ them?

. . .

Today:

4) How do you specify __Time Spans__?

5) How do you account for __Time Zones__?

. . .

\

```{r}
#| eval: false
library(tidyverse)  # includes "lubridate"
```


# Time Spans

## How old is Oski?

Oski, the official mascot for UC Berkeley, was debuted on September 27, 1941, during a
freshman rally at the Greek Theatre.

. . .

```{r}
#| output-location: fragment
a <- today()
b <- ymd("1941-09-27")
class(a)
```

. . .

```{r}
#| output-location: fragment
class(b)
```

. . .

```{r}
#| output-location: fragment
a - b
```


## Durations

Subtraction of two base R `Date` vectors -> `difftime` vector.

. . .

```{r}
#| output-location: fragment
class(a - b)
```

. . .

A `difftime` can be recorded in many different units, so better to use `lubridate::as.duration()` to always record it in [seconds]{.bold-hilit}.

. . .

```{r}
#| output-location: fragment
oski_age_sec <- as.duration(a - b)
class(oski_age_sec)
```

. . .

```{r}
#| output-location: fragment
oski_age_sec
```


## Constructing a duration

You can make a `duration` by coercing a `Date` object or using a constructor function.

. . .

:::: {.columns}
::: {.column .nonincremental width="50%"}
```{r}
dseconds(15)
```

```{r}
dminutes(10)
```

```{r}
dhours(c(12, 24))
```
:::

::: {.column .fragment .nonincremental width="50%"}
```{r}
ddays(0:3)
```

```{r}
dweeks(3)
```

```{r}
dyears(1)
```
:::
::::


## Some Wrinkles

. . .

```{r}
#| output-location: fragment
one_am <- ymd_hms("2026-03-08 01:00:00", tz = "America/Los_Angeles")
one_am
```

. . .

\

What should the following yield?

```{r}
#| output-location: fragment
one_am + ddays(1)
```

. . .

> Daylight savings time!


## Question

What is the answer to the following questions:

1) How long is a __day__?

2) How long is a __year__?


## Periods

In `"lubridate"`, a `period` is like a `duration` except it doesn’t have a fixed length in seconds. It tracks human units (day, week) that are responsive to irregularities.

. . .

:::: {.columns}
::: {.column .nonincremental width="50%"}
Example period constructors:

```{r}
seconds(15)
```

```{r}
minutes(10)
```

```{r}
hours(c(12, 24))
```
:::

::: {.column .fragment .nonincremental width="50%"}
\
```{r}
days(0:3)
```

```{r}
weeks(3)
```

```{r}
years(1)
```
:::
::::


## Working with periods

. . .

```{r}
#| output-location: fragment
one_am
```

. . .

```{r}
#| output-location: fragment
one_am + days(1)
```

. . .

\

What will the following yield?

```{r}
#| output-location: fragment
ymd("2024-01-01") + dyears(1)
```

. . .

```{r}
#| output-location: fragment
ymd("2024-01-01") + years(1) # accounts for leap year
```


## How old will Oski be ...

. . .

In a year?

```{r}
#| output-location: fragment
oski_age_sec
```

. . .

```{r}
#| output-location: fragment
oski_age_sec + dyears(1) # adding average seconds in a year
```

. . .

```{r}
#| output-location: fragment
as.period(oski_age_sec)
```

. . .

```{r}
#| output-location: fragment
as.period(oski_age_sec) + years(1)
```


## Time Spans recap

### Base R

- `difftime`: inconsistent units

\

### lubridate

- `duration`: fixed span in seconds

- `period`: variable span in human units

- `interval`: duration plus a starting point


# Time Zones

## {background-image="images/World_Time_Zones_Map.png" background-size="contain"}

## {background-image="images/alcan.png" background-size="contain"}

## {background-image="images/india.png" background-size="contain"}

## {background-image="images/china.png" background-size="contain"}


## UTC

Coordinated Universal Time (UTC) is primary global time standard. Successor to GMT.


### Offsets

Each time zone has a particular offset from UTC.

- London: +0
- New York: -5
- Kyoto: +9

::: {.fragment}

```{r}
#| output-location: fragment
Sys.timezone() # What R says my timezone is
```
:::


## Olson Name

```{r}
# names of Time Zones
OlsonNames()
```


## Changing Time

Two flavors

Change the instant in time.


```{r}
#| output-location: fragment
x1 <- ymd_hms("2024-06-01 12:00:00", tz = "America/New_York")
x1
```

. . .

```{r}
#| output-location: fragment
force_tz(x1, tzone = "Asia/Tokyo")
```

. . .

Keep the same instant but change the way it's printed:

```{r}
#| output-location: fragment
with_tz(x1, tzone = "Asia/Tokyo")
```
