---
title: "Lab 4"
subtitle: "Basics of Functions"
author: "STAT 133"
format: stat133slides-revealjs
chalkboard: true
---

## Agenda

- Review: Basics of functions
- If else statements, for loops
- Quarto Slides
    + Create your first slides
- QuartoPub
    + Sign up
- 3rd Quarto Publication
- Problem Set 3

## Review: Basics of Functions
- Functions are defined using the `function()` command
- Functions can take arguments (inputs) and return values (outputs)
- Example:
```{r}
#| eval: false
add <- function(x, y) {
  return(x + y)
}
result <- add(3, 5)
print(result)  # Output: 8
```

## If Else Statements
- Used for conditional execution of code blocks
- Syntax:
```{r}
#| eval: false
if (condition) {
  # code to execute if condition is TRUE
} else {
  # code to execute if condition is FALSE
}
```

## Example of If Else
```{r}
#| eval: false
check_number <- function(num) {
  if (num > 0) {
    return("Positive")
  } else if (num < 0) {
    return("Negative")
  } else {
    return("Zero")
  }
}
print(check_number(10))  # Output: "Positive"
print(check_number(-5))  # Output: "Negative"
print(check_number(0))   # Output: "Zero"
```

## For Loops
- Used to iterate over a sequence (like a vector or list)
- Syntax:
```{r}
#| eval: false
for (variable in sequence) {
  # code to execute for each element in sequence
}
```

## Example of For Loop
```{r}
#| eval: false
sum_vector <- function(vec) {
  total <- 0
  for (num in vec) {
    total <- total + num
  }
  return(total)
}
print(sum_vector(c(1, 2, 3, 4, 5)))  # Output: 15
```


## Quarto Slides
- Quarto allows you to create slides using markdown syntax
- Use `#` or `##` to separate slides
- Example:
```{.markdown code-line-numbers="false"}
# Slide Title Main Heading
Content of the slide goes here.
## Next Slide - Subheading
More content here.
```

## What can slides contain?
- Text
- Code chunks
- Figures

We use Quarto for slides in STAT 133!

## Like last week: Folder `stat133/`

```{.markdown code-line-numbers="false"}
  stat133/
    labs/
        lab1/
        lab2/
        lab3/
    psets/
        pset1/
        pset2/
```


## Terminal in IDE

1) Launch your IDE (Positron or RStudio)

2) Launch your [Terminal]{.bold-hilit} (next to the _Console_ tab)

3) `cd` to your `stat133/labs/` folder

4) create a folder for this lab: `mkdir lab4`

5) `cd` to `lab4`

. . .

```{.markdown filename="Terminal" code-line-numbers="false"}
# create an empty qmd file "demo-pub3.qmd"
# (to be used for your 2nd Quarto demo publication)
touch demo-pub3.qmd
```

## Open your `demo-pub3.qmd` document {.smaller}

Make sure your `qmd` file has the following content (or something similar):

```{.markdown filename="demo-pub2.qmd" code-line-numbers="false"}
---
title: "Demo Publication 3"
subtitle: "First Contact with Quarto Slides"
author: "Your Name"
format: revealjs
---
```

## Adding code:
Code is added the same way as in regular `qmd` files, using code chunks.

```{r}
#| eval: false
# Example code chunk
x <- 10
y <- 20
sum <- x + y
```

## Slide Themes

Example:
```{.markdown code-line-numbers="false"}
format: revealjs:
  theme: solarized
```
Explore more themes on the [Quarto Website](https://quarto.org/docs/presentations/revealjs/themes.html)

## Line highlighting
```{r}
#| eval: false
# Example function to calculate square
square <- function(x) {
  return(x^2)
}
```

## Adding Slide Backgrounds {.smaller}

Single color:
```{.markdown code-line-numbers="false"}
## Slide Title {background-color="aquamarine"}
```
Gradient:
```{.markdown code-line-numbers="false"}
## Slide Title {background-gradient="linear-gradient(to bottom, #283b95, #17b2c3)"}
## Slide Title {background-gradient="radial-gradient(#283b95, #17b2c3)"}
```

# Quarto Pub

## Requirements:
Your slides must contain:
- Title slide with a link to [Quarto](https://quarto.org/)
- A function, for loop, and if else statement
- A highlighted line of code
- A theme
- A slide background (color or gradient)
- An image

## Publishing Steps

Assume I have a file called `demo-pub3.qmd` opened in Positron. To publish my `qmd` file to my QuartoPub account I follow these steps:

1) Go to the __Terminal__ tab.
    + Make sure your working directory contains the qmd file.
2) Type command: `quarto publish demo-pub3.qmd`
3) Choose provider [Quarto Pub]{.bold-hilit} and press return (enter)
4) Publish with default account (linked to my email address)
5) Optionally: you can change the name of the document
6) Press return (enter)

. . .

Once published, there will be an associated link (URL), something like:
<https://username.quarto.pub/demo-pub-3>


## Submission to bCourses

Submit the URL (link) of your published document to the corresponding assignment in bCourses.

See [Assignments tab > Quarto Publications > Pub3]{.bold-hilit}


# Problem Set 3

## Pset 3

Available in bCourses:

See [Assignments tab > Problem Sets > PS3]{.bold-hilit}

You'll find 2 files:

- `ps3.html`: instructions

- `ps3.qmd`: template file to write your answers

. . .

In case of trouble: Go to Files tab, folder `problem-sets`, folder `ps3`, and download the `qmd` file.

Sometimes Safari blocks or denies you access. If this is the case you may want to use another browser (e.g. Chrome).


## Pset 3 Submission

- Submit your `qmd` and `html` files to bCourses.
- Use assignment problem set __ps3__
- Graded credit / no-credit
- Credit for evidence of earnest engagement (just don't submit a blank or empty template file)




