---
title: "Interactive Graphics"
subtitle: "Data Visualization III"
format: stat133slides-revealjs
revealjs-plugins:
  - drop
chalkboard: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE, cache = FALSE)
library(tidyverse)
library(plotly)
library(ggiraph)
```


## Agenda

. . .

Recap

. . .

\

Objects of class `"ggplot"`

\

. . .

Adding interactivity to ggplots with packages:

- `"plotly"`
- `"ggiraph"`


## Palmer Penguins {.smaller}

```{r}
library(palmerpenguins)
penguins
```


## 

What are the mappings, geometries and settings used here?

```{r}
#| echo: false
penguins |> 
  drop_na() |> 
  ggplot(aes(x = bill_length_mm, y = flipper_length_mm)) +
  geom_point(aes(color = species, size = body_mass_g),
  alpha = 0.5, shape = 18) +
  facet_wrap(~ sex) +
  labs(title = "Palmer Penguins",
       x = "Bill Length (mm)",
       y = "Flipper Length (mm)") +
  theme_bw()
```

```{r}
#| echo: false
countdown::countdown(1.5, bottom = 0)
```


## 

What are the mappings, geometries and settings used here?

```{r}
#| echo: false
penguins |> 
  drop_na() |> 
  ggplot(aes(x = species, y = flipper_length_mm)) +
  geom_boxplot(aes(fill = species), alpha = 0.3) +
  geom_jitter(width = 0.05, 
    aes(color = species),
    size = 2,
    shape = 21) +
  labs(title = "Palmer Penguins",
       x = "",
       y = "Flipper Length (mm)") +
  theme_bw() +
  theme(legend.position = "none")
```

```{r}
#| echo: false
countdown::countdown(1.5, bottom = 0)
```



# `"ggplot"` Objects


## {.smaller .nostretch}

:::: {.columns}
::: {.column width="50%"}
```{r}
#| eval: false
ggplot(
  data = penguins,
  mapping = aes(x = bill_length_mm,
                y = flipper_length_mm,
                color = species)) +
  geom_point(size = 2)
```
:::

::: {.column .fragment width="50%"}
```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 7
ggplot(
  data = penguins,
  mapping = aes(x = bill_length_mm,
                y = flipper_length_mm,
                color = species)) +
  geom_point(size = 2)
```
:::

::::

## {.smaller .nostretch}

:::: {.columns}
::: {.column width="50%"}
```{r}
#| eval: false
# we can store ggplot objects
gg <- ggplot(
  data = penguins,
  mapping = aes(x = bill_length_mm,
                y = flipper_length_mm,
                color = species)) +
  geom_point(size = 2)

gg
```
:::

::: {.column .fragment width="50%"}
```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 7
gg <- ggplot(
  data = penguins,
  mapping = aes(x = bill_length_mm,
                y = flipper_length_mm,
                color = species)) +
  geom_point(size = 2)

gg
```
:::

::::


## `"ggplot"` objects

. . .

```{r}
#| output-location: fragment
# we can store ggplot objects
gg <- ggplot(
  data = penguins,
  mapping = aes(x = bill_length_mm,
                 y = flipper_length_mm,
                 color = species)) +
  geom_point(size = 2)

class(gg)
```

. . .

```{r}
#| output-location: fragment
names(gg)
```


## `"ggplot"` objects

. . .

```{r}
#| output-location: fragment
#| fig-align: center
#| fig-width: 10
#| fig-height: 7
#| out-width: "65%"
# we can add layers to ggplot objects
gg + theme_minimal()
```


## 

```{r}
#| output-location: fragment
#| fig-align: center
#| fig-width: 10
#| fig-height: 7
#| out-width: "65%"
gg1 <- gg + theme_minimal()
gg2 <- gg1 + labs(title = "Bill Length and Flipper Length")
gg2
```

## `"ggplot"` objects

. . .

```{r}
#| eval: false
#| fig-align: center
#| fig-width: 10
#| fig-height: 7
#| out-width: "65%"
# we can save (export) ggplot graphics
# e.g. save to working directory
ggsave(filename = "scatterplot1.png", 
       plot = gg, 
       width = 9,
       height = 7, 
       units = "in")
```


# Example

## {.smaller .nostretch}

:::: {.columns}
::: {.column width="50%"}
```{r}
#| eval: false
p1 <- ggplot(penguins,
  aes(x = bill_length_mm,
      y = flipper_length_mm,
      color = species)) +
  geom_point(alpha = 0) +
  theme_minimal()

p1
```
:::

::: {.column .fragment width="50%"}
```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 7
p1 <- ggplot(penguins,
  aes(x = bill_length_mm,
      y = flipper_length_mm,
      color = species)) +
  geom_point(alpha = 0) +
  theme_minimal()

p1
```
:::

::::


## {.smaller .nostretch}

:::: {.columns}
::: {.column width="50%"}
```{r}
#| eval: false
p2 <- p1 +
  geom_point(
    data = filter(penguins, species == "Adelie"),
    aes(x = bill_length_mm,
        y = flipper_length_mm))

p2
```
:::

::: {.column width="50%"}
```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 7
p2 <- p1 +
  geom_point(
    data = filter(penguins, species == "Adelie"),
    aes(x = bill_length_mm,
        y = flipper_length_mm))

p2
```
:::

::::


## {.smaller .nostretch}

:::: {.columns}
::: {.column width="50%"}
```{r}
#| eval: false
p3 <- p1 +
  geom_point(
    data = filter(penguins, 
    species %in% c("Adelie", "Chinstrap")),
    aes(x = bill_length_mm,
        y = flipper_length_mm))

p3
```
:::

::: {.column width="50%"}
```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 7
p3 <- p1 +
  geom_point(
    data = filter(penguins, 
    species %in% c("Adelie", "Chinstrap")),
    aes(x = bill_length_mm,
        y = flipper_length_mm))

p3
```
:::

::::


## {.smaller .nostretch}

:::: {.columns}
::: {.column width="50%"}
```{r}
#| eval: false
p4 <- p1 +
  geom_point(
    data = penguins,
    aes(x = bill_length_mm,
        y = flipper_length_mm,
        color = species))

p4
```
:::

::: {.column width="50%"}
```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 7
p4 <- p1 +
  geom_point(
    data = penguins,
    aes(x = bill_length_mm,
        y = flipper_length_mm,
        color = species))

p4
```
:::

::::


# `plotly` and `ggiraph`

## Installation

```{r}
#| eval: false
# run this command on R's console
# (don't include this in a qmd file)
install.packages(c("plotly", "ggiraph"))
```


# `"plotly"`

## About Plotly

__Plotly__ is a software company that develops online data analytics and visualization tools, as well as scientific graphing libraries for R, Python, Matlab, etc.

\

. . .

In R, we have the package `"plotly"`.

\

. . .

_plotly_ comes with its own grammar of graphics syntax. Interestingly, we can use the `ggplotly()` function to make a ggplot graphic into a web-interactive graphic.


## {.smaller .nostretch auto-animate=true}

::: {.columns}
::: {.column width="50%"}
```{r}
#| eval: false
ggplot(data = penguins,
       mapping = aes(x = bill_length_mm, 
                     y = flipper_length_mm,
                     color = species)) +
  geom_point()
```
:::

::: {.column width="50%"}
```{r height="100%"}
#| echo: false
#| fig-width: 8
#| fig-height: 7
ggplot(data = penguins,
       mapping = aes(x = bill_length_mm, 
                     y = flipper_length_mm,
                     color = species)) +
  geom_point()
```
:::
:::


## {.smaller .nostretch auto-animate=true}

::: {.columns}
::: {.column width="50%"}
```{r}
#| eval: false
gg1 <- ggplot(data = penguins,
       mapping = aes(x = bill_length_mm, 
                     y = flipper_length_mm,
                     color = species)) +
  geom_point()

ggplotly(gg1)
```
:::

::: {.column width="50%"}
```{r height="100%"}
#| echo: false
#| fig-width: 8
#| fig-height: 7
gg1 <- ggplot(data = penguins,
       mapping = aes(x = bill_length_mm, 
                     y = flipper_length_mm,
                     color = species)) +
  geom_point()

ggplotly(gg1)
```
:::
:::


## Be careful {.smaller}

Not all ggplot graphical elements translate to plotly

::: {.columns}
::: {.column width="50%"}
```{r}
#| eval: false
gg2 <- ggplot(data = penguins,
       mapping = aes(x = bill_length_mm, 
                     y = flipper_length_mm,
                     color = species)) +
  geom_point() +
  labs(title = "Scatter plot", 
       subtitle = "Palmen Penguins Data")

ggplotly(gg2)
```
:::

::: {.column width="50%"}
```{r height="100%"}
#| echo: false
#| fig-width: 8
#| fig-height: 7
gg2 <- ggplot(data = penguins,
       mapping = aes(x = bill_length_mm, 
                     y = flipper_length_mm,
                     color = species)) +
  geom_point() +
  labs(title = "Scatter plot", 
       subtitle = "Palmen Penguins Data")

ggplotly(gg2)
```
:::
:::


# `"ggiraph"`

## About `"ggiraph"`

This package allows us to make ggplot2 graphics interactive following the same framework and syntax of adding layers to a graphic.

\

`"ggiraph"` provides sibling _geom_ functions, e.g.:

- `geom_point_interactive()`
- `geom_bar_interactive()`
- `geom_col_interactive()`
- `geom_histogram_interactive()`
- `geom_density_interactive()`
- `geom_boxplot_interactive()`


## {.smaller .nostretch auto-animate=true}

::: {.columns}
::: {.column width="50%"}
```{r}
#| eval: false
ggplot(data = penguins,
       mapping = aes(x = bill_length_mm, 
                     y = flipper_length_mm,
                     color = species)) +
  geom_point()
```
:::

::: {.column width="50%"}
```{r height="100%"}
#| echo: false
#| fig-width: 8
#| fig-height: 7
ggplot(data = penguins,
       mapping = aes(x = bill_length_mm, 
                     y = flipper_length_mm,
                     color = species)) +
  geom_point()
```
:::
:::


## {.smaller .nostretch auto-animate=true}

::: {.columns}
::: {.column width="50%"}
```{r}
#| eval: false
gg3 <- ggplot(data = penguins,
       mapping = aes(x = bill_length_mm, 
                     y = flipper_length_mm,
                     color = species)) +
  geom_point_interactive(aes(tooltip = sex))

girafe(ggobj = gg3)
```
:::

::: {.column width="50%"}
```{r height="100%"}
#| echo: false
gg3 <- ggplot(data = penguins,
       mapping = aes(x = bill_length_mm, 
                     y = flipper_length_mm,
                     color = species)) +
  geom_point_interactive(aes(tooltip = sex))

girafe(ggobj = gg3)
```
:::
:::


##

```{r}
gg4 <- ggplot(
  data = penguins,
  aes(x = bill_length_mm, 
      y = flipper_length_mm,
       color = species,
      tooltip = sex)) +
  geom_point_interactive()

ggiraph::girafe(ggobj = gg4)
```
