---
title: "Pub 8"
subtitle: "Cleaning Text Data and Visualizing Activity"
author: "Your Name"
format: html
embed-resources: true
---

**Part 1 - Extract Information with Regex**
The message column contains multiple pieces of information.

*Tasks:*

Use regex to extract the log level (ERROR, WARN, INFO).

Extract the user ID number.

Extract the event type (failed-login, login-success, etc.).

Create new columns: `level`, `user_id`, `event`

```{r}
# Extract Information with Regex
```

**Part 2 - Work with Dates and Times**
Convert the `timestamp` column to a date-time object. Then extract: `hour of day` and `date`. Create `hour` and `date` columns.

```{r}
# Work with Dates and Times

```

**Part 3 - Create Chart of Events by Hour**

Show how many events occurred during each hour.

x-axis: `hour`

y-axis: `number of events`

```{r}
# Create Chart of Events by Hour
```

**Part 4 - Brief Analysis**
Explain what hour had the most events happening or other interesting findings in 1-2 sentences in the second Tab of your dashboard. 