---
title: "Pub 8 Solutions"
subtitle: "Cleaning Text Data and Visualizing Activity"
author: "STAT 133"
format: html
embed-resources: true
---

**Part 1 - Extract Information with Regex**
The message column contains multiple pieces of information.

*Tasks:*

Use regex to extract the log level (ERROR, WARN, INFO).

Extract the user ID number.

Extract the event type (failed-login, login-success, etc.).

Create new columns: `level`, `user_id`, `event`

```{r}
# Extract Information with Regex

times_clean <- times_data |>
  mutate(
    level = str_extract(message, "ERROR|WARN|INFO"),
    user_id = str_extract(message, "(?<=user=)\\d+"),
    event = str_extract(message, "(failed-login|login-success|retry-login)")
  )
```

**Part 2 - Work with Dates and Times**
Convert the `timestamp` column to a date-time object. Then extract: `hour of day` and `date`. Create `hour` and `date` columns.

```{r}
# Work with Dates and Times

times_clean <- times_clean |>
  mutate(
    timestamp = mdy_hm(timestamp),
    hour = hour(timestamp),
    date = as_date(timestamp)
  )
```

**Part 3 - Create Chart of Events by Hour**

Show how many events occurred during each hour.

x-axis: `hour`

y-axis: `number of events`

```{r}
# Create Chart of Events by Hour
events_by_hour <- times_clean |>
  count(hour)

events_chart <- ggplot(events_by_hour, aes(x = hour, y = n)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Events by Hour",
    x = "Hour of Day",
    y = "Number of Events"
  )
```

**Part 4 - Brief Analysis**
Explain what hour had the most events happening or other interesting findings in 1-2 sentences in the second Tab of your dashboard: \ 
Most activity occurs at 14:00 and 15:00, each with two events recorded. The data suggests login activity is concentrated in the early afternoon, with fewer events later in the day.