---
title: "The Grammar of Graphics"
subtitle: "Data Visualization II"
format: stat133slides-revealjs
revealjs-plugins:
  - drop
chalkboard: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE)
library(tidyverse)
library(patchwork)
```


# Recap

## Palmer Penguins {.smaller}

```{r}
library(palmerpenguins)
penguins
```


## `ggplot()` syntax

. . .

`ggplot()` builds a plot layer by layer, each one added on top of one another with `+` (not `|>`).

- `ggplot()` creates canvas
- `aes()` creates mappings, called inside `ggplot()` or a `geom()`
- `geom_()` puts down marks using declared geometry

. . .

- Usually you specify the `data` in `ggplot()`
- the `data` must be a dataframe or tibble


## {.smaller .nostretch auto-animate=true}

::: {.columns}
::: {.column width="50%"}
```{r}
#| eval: false
ggplot(data = penguins,
       mapping = aes(x = bill_length_mm, 
                     y = flipper_length_mm))
```
:::

::: {.column width="50%"}
```{r height="100%"}
#| echo: false
#| fig-width: 8
#| fig-height: 7
ggplot(data = penguins,
       mapping = aes(x = bill_length_mm, y = flipper_length_mm))
```
:::
:::

## {.smaller .nostretch auto-animate=true}

::: {.columns}
::: {.column width="50%"}
```{r}
#| eval: false
ggplot(data = penguins,
       mapping = aes(x = bill_length_mm, 
                     y = flipper_length_mm)) +
  geom_point()
```
:::

::: {.column width="50%"}
```{r height="100%"}
#| echo: false
#| fig-width: 8
#| fig-height: 7
ggplot(data = penguins,
       mapping = aes(x = bill_length_mm, y = flipper_length_mm)) +
  geom_point()
```
:::
:::

## {.smaller .nostretch auto-animate=true}

::: {.columns}
::: {.column width="50%"}
```{r}
#| eval: false
ggplot(data = penguins,
       mapping = aes(x = bill_length_mm, 
                     y = flipper_length_mm,
                     color = species)) +
  geom_point()
```
:::

::: {.column width="50%"}
```{r height="100%"}
#| echo: false
#| fig-width: 8
#| fig-height: 7
ggplot(data = penguins,
       mapping = aes(x = bill_length_mm, 
                     y = flipper_length_mm,
                     color = species)) +
  geom_point()
```
:::
:::


# More on Mappings

## Recap

- An [aesthetic mapping]{.hilit} links a variable in the data to a visual attribute.
- `ggplot()` and `geom_()` have a `mapping` argument.
- We use the `aes()` to specify such mappings, e.g.
    + `ggplot(..., aes(x = species))`
    + `geom_bar(aes(x = species))`


## Global -vs- Local Mappings

We distinguish 2 types of aesthetic mappings:

- [Global Mappings]{.hilit}: those specified inside `ggplot()`

\

- [Local Mappings]{.hilit}: those specified inside `geom_()` functions


##  {.smaller .nostretch}

::::: columns
::: {.column width="50%"}
```{r}
#| output-location: fragment
#| fig-width: 8
#| fig-height: 7
# global mappings
ggplot(data = penguins,
       aes(x = bill_length_mm, 
           y = flipper_length_mm)) +
  geom_point()
```
:::

::: {.column .fragment width="50%"}
```{r}
#| output-location: fragment
#| fig-width: 8
#| fig-height: 7
# local mappings
ggplot(data = penguins) +
  geom_point(
    aes(x = bill_length_mm, 
        y = flipper_length_mm))  
```
:::
:::::


##  {.smaller .nostretch}

::::: columns
::: {.column width="50%"}
```{r}
#| output-location: fragment
#| fig-width: 8
#| fig-height: 7
# global mappings
ggplot(data = penguins,
       aes(x = bill_length_mm, 
           y = flipper_length_mm)) +
  geom_point() +
  geom_smooth(method = "lm")
```
:::

::: {.column .fragment width="50%"}
```{r height="100%"}
#| output-location: fragment
#| fig-width: 8
#| fig-height: 7
# local mappings (be careful!!!)
ggplot(data = penguins) +
  geom_point(
    aes(x = bill_length_mm, 
        y = flipper_length_mm))  +
  geom_smooth(method = "lm")
```
:::
:::::


##  {.smaller .nostretch}

::::: columns
::: {.column width="50%"}
```{r}
#| fig-width: 8
#| fig-height: 7
# global mappings
ggplot(data = penguins,
       aes(x = bill_length_mm, 
           y = flipper_length_mm)) +
  geom_point() +
  geom_smooth(method = "lm")
```
:::

::: {.column .fragment width="50%"}
```{r height="100%"}
#| output-location: fragment
#| fig-width: 8
#| fig-height: 7
# local mappings (be careful!!!)
ggplot(data = penguins) +
  geom_point(aes(x = bill_length_mm, 
                 y = flipper_length_mm))  +
  geom_smooth(aes(x = bill_length_mm, 
                  y = flipper_length_mm),
              method = "lm")
```
:::
:::::


##  {.smaller .nostretch}

::::: columns
::: {.column width="50%"}
```{r}
#| output-location: fragment
#| fig-width: 8
#| fig-height: 7
# global mappings
ggplot(data = penguins,
       aes(x = bill_length_mm, 
           y = flipper_length_mm,
           color = species)) +
  geom_point()
```
:::

::: {.column .fragment width="50%"}
```{r}
#| output-location: fragment
#| fig-width: 8
#| fig-height: 7
# local mappings
ggplot(data = penguins) +
  geom_point(
    aes(x = bill_length_mm, 
        y = flipper_length_mm,
        color = species))  
```
:::
:::::


##  {.smaller .nostretch}

::::: columns
::: {.column width="50%"}
```{r}
#| output-location: fragment
#| fig-width: 8
#| fig-height: 7
# global mappings
ggplot(data = penguins,
       aes(x = bill_length_mm, 
           y = flipper_length_mm,
           color = species)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm")
```
:::

::: {.column .fragment width="50%"}
```{r height="100%"}
#| output-location: fragment
#| fig-width: 8
#| fig-height: 7
# global and local mappings
ggplot(data = penguins,
       aes(x = bill_length_mm, 
           y = flipper_length_mm)) +
  geom_point(size=3, aes(color = species)) +
  geom_smooth(method = "lm",
              color = "black")
```
:::
:::::


##  {.smaller .nostretch}

::::: columns
::: {.column width="50%"}
```{r}
#| output-location: fragment
#| fig-width: 8
#| fig-height: 7
# global mappings
ggplot(data = penguins,
       aes(x = bill_length_mm, 
           y = flipper_length_mm,
           color = species)) +
  geom_smooth(method = "lm") +
  geom_point(size = 3)
```
:::

::: {.column .fragment width="50%"}
```{r height="100%"}
#| output-location: fragment
#| fig-width: 8
#| fig-height: 7
# global and local mappings
ggplot(data = penguins,
       aes(x = bill_length_mm, 
           y = flipper_length_mm)) +
  geom_smooth(method = "lm",
              color = "black") +
  geom_point(size=3, aes(color = species))
```
:::
:::::


# More Graphical Elements

## Additional graphical elements

- Labels
    + Title
    + Subtitle
    + Axis labels
- Annotations
- Themes
- Facets
- Scales


## Labels

`labs()` allows you to add

```r
labs(
  title = "...",
  subtitle = "...",
  x = "...",
  y = "..."
)
```


## {.smaller .nostretch auto-animate=true}

::: {.columns}
::: {.column width="50%"}
```{r}
#| eval: false
ggplot(data = penguins,
       mapping = aes(x = bill_length_mm, 
                     y = flipper_length_mm,
                     color = species)) +
  geom_point()
```
:::

::: {.column width="50%"}
```{r height="100%"}
#| echo: false
#| fig-width: 8
#| fig-height: 7
ggplot(data = penguins,
       mapping = aes(x = bill_length_mm, 
                     y = flipper_length_mm,
                     color = species)) +
  geom_point()
```
:::
:::


## {.smaller .nostretch auto-animate=true}

::: {.columns}
::: {.column width="50%"}
```{r}
#| eval: false
ggplot(data = penguins,
       mapping = aes(x = bill_length_mm, 
                     y = flipper_length_mm,
                     color = species)) +
  geom_point() +
  labs(
    title = "Bill and flipper length",
    subtitle = "Positive relation",
    x = "Bill length (mm)",
    y = "Flipper length (mm)")
```
:::

::: {.column width="50%"}
```{r height="100%"}
#| echo: false
#| fig-width: 8
#| fig-height: 7
ggplot(data = penguins,
       mapping = aes(x = bill_length_mm, 
                     y = flipper_length_mm,
                     color = species)) +
  geom_point() +
  labs(
    title = "Bill and flipper length",
    subtitle = "Positive relation",
    x = "Bill length (mm)",
    y = "Flipper length (mm)")
```
:::
:::


# Facets

## Facets

- `facet_wrap()` wraps a 1d sequence of panels into 2d

- `facet_grid()` forms a matrix of panels defined by row and column faceting variables

    + `facet_grid(rows = vars(...))`
    + `facet_grid(cols = vars(...))`
    + `facet_grid(row_var ~ col_var)`


## {.smaller .nostretch auto-animate=true}

```{r}
#| out-width: "95%"
#| output-location: fragment
ggplot(data = penguins,
       mapping = aes(x = bill_length_mm, y = flipper_length_mm, color = species)) +
  geom_point() +
  facet_wrap(~ species)
```


## {.smaller .nostretch auto-animate=true}

```{r}
#| out-width: "95%"
#| output-location: fragment
ggplot(data = penguins,
       mapping = aes(x = bill_length_mm, y = flipper_length_mm, color = species)) +
  geom_point() +
  facet_wrap(~ species, nrow = 2)
```


## {.smaller .nostretch auto-animate=true}

```{r}
#| out-width: "95%"
#| output-location: fragment
ggplot(data = penguins,
       mapping = aes(x = bill_length_mm, y = flipper_length_mm, color = species)) +
  geom_point() +
  facet_grid(. ~ species)  # facet_grid(cols = vars(species))
```


## {.smaller .nostretch auto-animate=true}

```{r}
#| out-width: "95%"
#| output-location: fragment
ggplot(data = penguins,
       mapping = aes(x = bill_length_mm, y = flipper_length_mm, color = species)) +
  geom_point() +
  facet_grid(species ~ .)  # facet_grid(rows = vars(species))
```


## {.smaller .nostretch auto-animate=true}

```{r}
#| out-width: "95%"
#| output-location: fragment
ggplot(data = penguins,
       mapping = aes(x = bill_length_mm, y = flipper_length_mm, color = species)) +
  geom_point() +
  facet_grid(species ~ island)
```


# Themes

## Theme Functions

- `theme_gray()`
- `theme_bw()`
- `theme_minimal()`
- `theme_classic()`
- `theme_light()`

. . .

There's also the package `"ggthemes"`

<https://yutannihilation.github.io/allYourFigureAreBelongToUs/ggthemes/>


## {.smaller .nostretch auto-animate=true}

::: {.columns}
::: {.column width="50%"}
```{r}
#| eval: false
ggplot(data = penguins,
       mapping = aes(x = bill_length_mm, 
                     y = flipper_length_mm,
                     color = species)) +
  geom_point()
```
:::

::: {.column width="50%"}
```{r height="100%"}
#| echo: false
#| fig-width: 8
#| fig-height: 7
ggplot(data = penguins,
       mapping = aes(x = bill_length_mm, 
                     y = flipper_length_mm,
                     color = species)) +
  geom_point()
```
:::
:::


## {.smaller .nostretch auto-animate=true}

::: {.columns}
::: {.column width="50%"}
```{r}
#| eval: false
ggplot(data = penguins,
       mapping = aes(x = bill_length_mm, 
                     y = flipper_length_mm,
                     color = species)) +
  geom_point() +
  theme_classic()
```
:::

::: {.column width="50%"}
```{r height="100%"}
#| echo: false
#| fig-width: 8
#| fig-height: 7
ggplot(data = penguins,
       mapping = aes(x = bill_length_mm, 
                     y = flipper_length_mm,
                     color = species)) +
  geom_point() +
  theme_classic()
```
:::
:::


## {.smaller .nostretch auto-animate=true}

::: {.columns}
::: {.column width="50%"}
```{r}
#| eval: false
ggplot(data = penguins,
       mapping = aes(x = bill_length_mm, 
                     y = flipper_length_mm,
                     color = species)) +
  geom_point() +
  theme_bw()
```
:::

::: {.column width="50%"}
```{r height="100%"}
#| echo: false
#| fig-width: 8
#| fig-height: 7
ggplot(data = penguins,
       mapping = aes(x = bill_length_mm, 
                     y = flipper_length_mm,
                     color = species)) +
  geom_point() +
  theme_bw()
```
:::
:::


## {.smaller .nostretch auto-animate=true}

::: {.columns}
::: {.column width="50%"}
```{r}
#| eval: false
ggplot(data = penguins,
       mapping = aes(x = bill_length_mm, 
                     y = flipper_length_mm,
                     color = species)) +
  geom_point() +
  theme_minimal()
```
:::

::: {.column width="50%"}
```{r height="100%"}
#| echo: false
#| fig-width: 8
#| fig-height: 7
ggplot(data = penguins,
       mapping = aes(x = bill_length_mm, 
                     y = flipper_length_mm,
                     color = species)) +
  geom_point() +
  theme_minimal()
```
:::
:::


# Scales

## Scale Functions

- `scale_x_continuous()`
- `scale_y_continuous()`
- `scale_color_manual()`
- `scale_fill_manual()`
- ETC


## {.smaller .nostretch auto-animate=true}

::: {.columns}
::: {.column width="50%"}
```{r}
#| eval: false
ggplot(data = penguins,
       mapping = aes(x = bill_length_mm, 
                     y = flipper_length_mm,
                     color = species)) +
  geom_point(size = 3)
```
:::

::: {.column width="50%"}
```{r height="100%"}
#| echo: false
#| fig-width: 8
#| fig-height: 7
ggplot(data = penguins,
       mapping = aes(x = bill_length_mm, 
                     y = flipper_length_mm,
                     color = species)) +
  geom_point(size = 3)
```
:::
:::


## {.smaller .nostretch auto-animate=true}

::: {.columns}
::: {.column width="50%"}
```{r}
#| eval: false
ggplot(data = penguins,
       mapping = aes(x = bill_length_mm, 
                     y = flipper_length_mm,
                     color = species)) +
  geom_point(size = 3) +
  scale_x_continuous(limits = c(0, 60))
```
:::

::: {.column width="50%"}
```{r height="100%"}
#| echo: false
#| fig-width: 8
#| fig-height: 7
ggplot(data = penguins,
       mapping = aes(x = bill_length_mm, 
                     y = flipper_length_mm,
                     color = species)) +
  geom_point(size = 3) +
  scale_x_continuous(limits = c(0, 60))
```
:::
:::

## {.smaller .nostretch auto-animate=true}

::: {.columns}
::: {.column width="50%"}
```{r}
#| eval: false
ggplot(data = penguins,
       mapping = aes(x = bill_length_mm, 
                     y = flipper_length_mm,
                     color = species)) +
  geom_point(size = 3) +
  scale_x_continuous(limits = c(0, 60)) +
  scale_y_continuous(limits = c(160, 240))
```
:::

::: {.column width="50%"}
```{r height="100%"}
#| echo: false
#| fig-width: 8
#| fig-height: 7
ggplot(data = penguins,
       mapping = aes(x = bill_length_mm, 
                     y = flipper_length_mm,
                     color = species)) +
  geom_point(size = 3) +
  scale_x_continuous(limits = c(0, 60)) +
  scale_y_continuous(limits = c(160, 240))
```
:::
:::


## {.smaller .nostretch auto-animate=true}

::: {.columns}
::: {.column width="50%"}
```{r}
#| eval: false
ggplot(data = penguins,
       mapping = aes(x = bill_length_mm, 
                     y = flipper_length_mm,
                     color = species)) +
  geom_point(size = 3)
```
:::

::: {.column width="50%"}
```{r height="100%"}
#| echo: false
#| fig-width: 8
#| fig-height: 7
ggplot(data = penguins,
       mapping = aes(x = bill_length_mm, 
                     y = flipper_length_mm,
                     color = species)) +
  geom_point(size = 3)
```
:::
:::


## {.smaller .nostretch auto-animate=true}

::: {.columns}
::: {.column width="50%"}
```{r}
#| eval: false
ggplot(data = penguins,
       mapping = aes(x = bill_length_mm, 
                     y = flipper_length_mm,
                     color = species)) +
  geom_point(size = 3) +
  scale_color_manual(
    values = c("orange", "magenta", "turquoise"))
```
:::

::: {.column width="50%"}
```{r height="100%"}
#| echo: false
#| fig-width: 8
#| fig-height: 7
ggplot(data = penguins,
       mapping = aes(x = bill_length_mm, 
                     y = flipper_length_mm,
                     color = species)) +
  geom_point(size = 3) +
  scale_color_manual(
    values = c("orange", "magenta", "turquoise"))
```
:::
:::


# Bonus Example

## {.smaller .nostretch}

```{r}
#| output-location: fragment
#| code-line-numbers: "1|1-2|1-3|1-4|1-6"
big_males <- penguins |> 
  filter(sex == "male") |> 
  group_by(species) |> 
  slice_max(body_mass_g)

big_males
```


## {.smaller .nostretch}

:::: {.columns}
::: {.column width="50%"}
```{r}
#| eval: false
#| output-location: fragment
ggplot() +
  geom_point(
    data = filter(penguins, sex == "male"),
    aes(x = bill_length_mm,
        y = flipper_length_mm,
        color = species),
    size = 0.5, alpha = 0.5) +
  geom_point(
    data = big_males,
    aes(x = bill_length_mm,
        y = flipper_length_mm,
        color = species),
    size = 4) +
  theme_minimal()
```
:::

::: {.column .fragment width="50%"}
```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 7
ggplot() +
  geom_point(
    data = filter(penguins, sex == "male"),
    aes(x = bill_length_mm,
        y = flipper_length_mm,
        color = species),
    size = 2, alpha = 0.5) +
  geom_point(
    data = big_males,
    aes(x = bill_length_mm,
        y = flipper_length_mm,
        color = species),
    size = 4) +
  theme_minimal()
```
:::

::::


##

Because the mappings for `x`, `y` and `color` are the same in both calls to `geom_point()`

::: {.nonincremental}
- `x = bill_length_mm`
- `y = flipper_length_mm`
- `color = species`
:::

we can simplify the command in the previous slides as follows (next slide)

## {.smaller .nostretch}

:::: {.columns}
::: {.column width="50%"}
```{r}
#| eval: false
#| output-location: fragment
ggplot(
  data = filter(penguins, sex == "male"),
  aes(x = bill_length_mm,
      y = flipper_length_mm,
      color = species)) +
  geom_point(size = 2, alpha = 0.5) +
  geom_point(
    data = big_males,
    size = 4) +
  theme_minimal()
```
:::

::: {.column .fragment width="50%"}
```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 7
ggplot(
  data = filter(penguins, sex == "male"),
  aes(x = bill_length_mm,
      y = flipper_length_mm,
      color = species)) +
  geom_point(size = 2, alpha = 0.5) +
  geom_point(
    data = big_males,
    size = 4) +
  theme_minimal()
```
:::

::::