---
title: "The Grammar of Graphics"
subtitle: "Data Visualization II"
format: stat133slides-revealjs
chalkboard: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE)
library(tidyverse)
library(patchwork)
```


# Recap

## Palmer Penguins {.smaller}

```{r}
library(palmerpenguins)
penguins
```


## `ggplot()` syntax

. . .

`ggplot()` builds a plot layer by layer, each one added on top of one another with `+` (not `|>`).

- `ggplot()` creates canvas
- `aes()` creates mappings, called inside `ggplot()` or a `geom()`
- `geom_()` puts down marks using declared geometry

. . .

- Usually you specify the `data` in `ggplot()`
- the `data` must be a dataframe or tibble


## {.smaller .nostretch auto-animate=true}

::: {.columns}
::: {.column width="50%"}
```{r}
#| eval: false
ggplot(data = penguins,
       mapping = aes(x = bill_length_mm, 
                     y = flipper_length_mm))
```
:::

::: {.column width="50%"}
```{r height="100%"}
#| echo: false
#| fig-width: 8
#| fig-height: 7
ggplot(data = penguins,
       mapping = aes(x = bill_length_mm, y = flipper_length_mm))
```
:::
:::

## {.smaller .nostretch auto-animate=true}

::: {.columns}
::: {.column width="50%"}
```{r}
#| eval: false
ggplot(data = penguins,
       mapping = aes(x = bill_length_mm, 
                     y = flipper_length_mm)) +
  geom_point()
```
:::

::: {.column width="50%"}
```{r height="100%"}
#| echo: false
#| fig-width: 8
#| fig-height: 7
ggplot(data = penguins,
       mapping = aes(x = bill_length_mm, y = flipper_length_mm)) +
  geom_point()
```
:::
:::

## {.smaller .nostretch auto-animate=true}

::: {.columns}
::: {.column width="50%"}
```{r}
#| eval: false
ggplot(data = penguins,
       mapping = aes(x = bill_length_mm, 
                     y = flipper_length_mm,
                     color = species)) +
  geom_point()
```
:::

::: {.column width="50%"}
```{r height="100%"}
#| echo: false
#| fig-width: 8
#| fig-height: 7
ggplot(data = penguins,
       mapping = aes(x = bill_length_mm, 
                     y = flipper_length_mm,
                     color = species)) +
  geom_point()
```
:::
:::


# More Graphical Elements

## Additional graphical elements

- Labels
    + Title
    + Subtitle
    + Axis labels
- Annotations
- Themes
- Scales
- Facets


## Labels

`labs()` allows you to add

```r
labs(
  title = "...",
  subtitle = "...",
  x = "...",
  y = "..."
)
```


## {.smaller .nostretch auto-animate=true}

::: {.columns}
::: {.column width="50%"}
```{r}
#| eval: false
ggplot(data = penguins,
       mapping = aes(x = bill_length_mm, 
                     y = flipper_length_mm,
                     color = species)) +
  geom_point()
```
:::

::: {.column width="50%"}
```{r height="100%"}
#| echo: false
#| fig-width: 8
#| fig-height: 7
ggplot(data = penguins,
       mapping = aes(x = bill_length_mm, 
                     y = flipper_length_mm,
                     color = species)) +
  geom_point()
```
:::
:::


## {.smaller .nostretch auto-animate=true}

::: {.columns}
::: {.column width="50%"}
```{r}
#| eval: false
ggplot(data = penguins,
       mapping = aes(x = bill_length_mm, 
                     y = flipper_length_mm,
                     color = species)) +
  geom_point() +
  labs(
    title = "Bill and flipper length",
    subtitle = "Positive relation",
    x = "Bill length (mm)",
    y = "Flipper length (mm)")
```
:::

::: {.column width="50%"}
```{r height="100%"}
#| echo: false
#| fig-width: 8
#| fig-height: 7
ggplot(data = penguins,
       mapping = aes(x = bill_length_mm, 
                     y = flipper_length_mm,
                     color = species)) +
  geom_point() +
  labs(
    title = "Bill and flipper length",
    subtitle = "Positive relation",
    x = "Bill length (mm)",
    y = "Flipper length (mm)")
```
:::
:::


# Facets

## Facets

- `facet_wrap()` wraps a 1d sequence of panels into 2d

- `facet_grid()` forms a matrix of panels defined by row and column faceting variables

    + `facet_grid(rows = vars(...))`
    + `facet_grid(cols = vars(...))`
    + `facet_grid(row_var ~ col_var)`


## {.smaller .nostretch auto-animate=true}

```{r}
#| out-width: "95%"
#| output-location: fragment
ggplot(data = penguins,
       mapping = aes(x = bill_length_mm, y = flipper_length_mm, color = species)) +
  geom_point() +
  facet_wrap(~ species)
```


## {.smaller .nostretch auto-animate=true}

```{r}
#| out-width: "95%"
#| output-location: fragment
ggplot(data = penguins,
       mapping = aes(x = bill_length_mm, y = flipper_length_mm, color = species)) +
  geom_point() +
  facet_grid(. ~ species)
```


## {.smaller .nostretch auto-animate=true}

```{r}
#| out-width: "95%"
#| output-location: fragment
ggplot(data = penguins,
       mapping = aes(x = bill_length_mm, y = flipper_length_mm, color = species)) +
  geom_point() +
  facet_grid(species ~ .)
```


# Themes

## Theme Functions

- `theme_gray()`
- `theme_bw()`
- `theme_minimal()`
- `theme_classic()`
- `theme_light()`

